Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-01-28 manuscript/manuscript.qmd
Study 2

We next examine if advisers believe that flattering advice will improve
how advisees perceive them, and whether this perception is true. We do this
by manipulating the incentives for advisers, who receive a bonus either based
on how they are evaluated by an advisee, or a bonus for the advisee's accuracy.

We conducted a three-stage experiment in which we experimentally
manipulated the incentives for advisers. We began by inviting a group of
advisees to upload photos of themselves ("selfies") and informed them
that they would be rated on their attractiveness. We grouped them with
nine other participants of the same gender and recruited participants of
the opposing gender to rank them from most to least attractive and to
provide advice. Specifically, we asked them to advise the participant
they ranked as the 7th most attractive (4th least attractive) on what
rank they should bet they were ranked by a larger group of raters.
Advisers were randomly assigned to two treatments, receiving a bonus
payment either if the advisee guessed their rank accurately or if the
advisee evaluated the adviser as likeable as measured by an
unincentivized scale response. We hypothesize that advisers who want to
be liked by advisees will recommend that they bet on a lower rank, i.e.
that they are more attractive.

@@

We recruited 300 participants from Prolific and, after asking
demographic questions, invited them to upload photos of themselves
(selfies) to be rated by other participants on attractiveness.
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women agreed to do so and uploaded pictures that adhered to our
instructions (e.g., did not include other people). We selected the first
100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years). Participants
were informed that their selfies would be randomly grouped with those of
nine other participants of their gender and ranked in terms of
attractiveness by a group of new Prolific participants of the opposite
gender. Lastly, they guessed their rank (unincentivized).

Next, we recruited a new, gender-balanced sample for the role of
advisers. `r scales::comma(dim(study2_time2)[1])` participants from
Prolific ($M_{\text{Age}}$ = `r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female) started
by providing demographic information, then were matched to a group of
the opposite gender. They then ranked participants from most to least
attractive using a drop-down menu next to each picture. Because of a
limitation with the survey software, participants were able to select
the same rank for multiple participants, rather than rank them uniquely.
We remove 115 participants who did not follow instructions and provide a
unique ranking.

After submitting their ratings, they saw the photo of the participant
they had ranked as the 7th most attractive (i.e., the 4th least
attractive). They were reminded of the rank they had given to that
person and informed that this participant would be invited back and
could earn a \$1 bonus if they guessed their rank correctly. The rank
was determined by the aggregate ratings of all participants who had
ranked this group. Their task was to give advice to this participant
about the rank they should bet on based on having observed all ten
selfies and their own assessment. We randomly assigned participants to
one of two incentivization schemes. In the "Accuracy" treatment, they
received a bonus identical to the advisee: \$1 if they guessed their
rank correctly. In the "Likeability" treatment, we informed them that
they would be rated by the advisee on a 5-point Likert scale for how
likeable they thought they were. Each point on the scale would translate
to a bonus of 20 cents. They then selected a rank that they would
recommend the advisee to bet on.

Finally, we invited participants from Stage 1 and were ranked as the 7th
attractive by at least one adviser (so that they received advice) back
for the follow-up survey. Following our preregistration, we kept the
survey open for 7 days. In total,
`r scales::comma(dim(study2_time3)[1])` participants
(`r scales::comma(dim(study2_time3[study2_time3$gender == 'Male',])[1])`
men,
`r scales::comma(dim(study2_time3[study2_time3$gender == 'Female',])[1])`
women) returned. They were reminded of the selfie they uploaded in Stage
1 and informed that a group of 10 selfies, including theirs, had been
rated by other participants from Prolific. They then saw the advice from
a randomly selected adviser and made their estimate with a \$1 incentive
for guessing accurately. They then saw the rank they had been advised to
bet on one more time and were asked to rate the adviser's likability,
warmth, friendliness, good-naturedness, trustworthiness, and sincerity
on 5-point Likert scales (adapted from @Fiske2007).

@@

```{r}
#| label: fig-study2advice
#| fig-cap: "Advisers were asked to suggest what rank to bet on for someone they ranked as the 7th most attractive out of ten people. When incentivized for accuracy, most of them advised betting on 7, which was based on their evaluation of the recipient's attractiveness. However, when incentivized for likability, many participants gave flattering advice by suggesting betting on higher ranks that imply a higher evaluation of attractiveness."

@@

    # Theme adjustments
    theme_bw() +
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(colour = "black"),
          legend.position = "bottom",
          legend.title = element_blank(),
          legend.key.size = unit(1.2, "lines"),
          legend.text = element_text(size = 12),
          plot.title = element_text(hjust = 0.5, size = 20),
          axis.text.x = element_text(size = 14),  # X-axis text size
          axis.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 16, face = "bold"),  # X-axis title size
          axis.title.y = element_text(size = 16, face = "bold"))

We begin by examining the prior beliefs of advisees who uploaded their
selfies. On average, men guessed they ranked
`r round(mean(study2_time1[study2_time1$gender == "Male",]$Guess),2)` in
their group of 10 and women guessed that they ranked
`r round(mean(study2_time1[study2_time1$gender == "Female",]$Guess),2)`.
Participants overall underestimate their attractiveness relative to the
benchmark average of 5.5. Moreover, women do so more than men
`r papaja::apa_print(t.test(Guess ~ gender, data = study2_time1, var.equal = T))$statistic`.
Notably, people's self-perceptions correlated strongly with the
aggregate ratings of the advisers
(`r round(cor(study2_finalRank$Guess, study2_finalRank$finalRank),2)`,
`r papaja::apa_print(cor.test(study2_finalRank$Guess, study2_finalRank$finalRank))$statistic`).
However, there was substantial heterogeneity in perceptions of
attractiveness. Of the 200 participants,
`r length(unique(study2_time2$Target))` were ranked as 7th most
attractive by at least one adviser. On average, men in this subset
estimated they were ranked
`r round(mean(TargetsUnique[TargetsUnique$gender == 'Male',]$Guess),2)`th
and women estimated they were ranked
`r round(mean(TargetsUnique[TargetsUnique$gender == 'Female',]$Guess),2)`th
(`r papaja::apa_print(t.test(Guess ~ gender, data = TargetsUnique, var.equal = T))$statistic`).

Next, we turn our attention to the advisers (@fig-study2advice). In the
Accuracy condition, those uploading selfies were advised to bet on rank
`r round(mean(study2_time2[study2_time2$Condition == "Accurate",]$Advice),2)`.
Notably, this is significantly more attractive than the 7th rank those
advisers had themselves guessed just on the prior screen
(`r papaja::apa_print(t.test(study2_time2[study2_time2$Condition == "Accurate",]$Advice, mu = 7, var.equal = T))$statistic`).
This suggests that even when incentivized for accuracy, participants
offered flattering advice.[^1] Importantly, and as predicted, we find
that advisers in the Likeable treatment recommend betting on a lower
rank, communicating that they think the participant in the selfie is
more attractive
(`r round(mean(study2_time2[study2_time2$Condition == "Likable",]$Advice),2)`,
`r papaja::apa_print(t.test(Advice ~ Condition, data = study2_time2, var.equal = T))$statistic`).
This suggests that advisers inferred that advising someone that they
were more attractive would make the adviser appear more likeable and
therefore provided advice that communicated a more favorable impression
of the participants' attractiveness. The distribution shown in
@fig-study2advice shows that participants do not simply tell
participants that they are the most attractive person in the group. They
may infer that flattering advise needs to be somewhat realistic to be
believable. We return to this in the general discussion.

[^1]: This shading could be due to concerns of avoiding disappointment.
    However, it could also be that advisers are uncertain about the
    rankings they had given and make a recommendation that combines
    their own belief with a uniform prior.

@@

# Create the plot
ggplot(study2_long, aes(x = Advice, y = Value, color = Evaluation)) +  
  # Add points for each observation
  geom_point(size = 2, alpha = 0.7) +

@@

```{r}
#| label: fig-study2ratingByGroup
#| fig-cap: "After participants received advice that suggested betting on a rank that was lower than, equal to, or higher than their self-expectations, they evaluated advisors based on their likability and warmth. Participants rated advisors who gave flattering advice and delivered a high evaluation of their attractiveness as more likable and warm than those whose advice implied a lower evaluation of their attractiveness."

@@

# Create the heatmap
ggplot(melted_data, aes(x = Advice, y = Condition, fill = Rating)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "grey80", high = "grey30", name = "Rating") +  # Grey gradient from light to dark
  geom_text(aes(label = sprintf("%.2f", Rating)), color = "white", size = 8) +
  theme_minimal() +
  labs(x = "Attractiveness (Advice Relative to Self-Expectations)", y = "Evaluation") +
  theme(
    axis.text.x = element_text(size = 12),  # X-axis text size
          axis.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 12),  # X-axis title size
          axis.title.y = element_text(size = 12))

@@

As seen in @fig-study2rating, we found that advisees who suggested that
the advisee was more attractive (lower rank) were indeed rated as more
likeable and warm (b = -0.109, p \< .05; b = -0.133, p \< .001,
respectively; See Columns 1 and 2 of @tbl-study2regs). When using their
self-expectations as a reference point, advisors who suggest a rank
lower than that point are perceived as less likable and warm (see
@fig-study2ratingByGroup). Interestingly, these benefits are not at the
cost of being hypocritical; advisers who recommend a more favorable rank
are viewed as no less trustworthy (Column 3 of @tbl-study2regs).

We were not powered to do a comparison across the two experimental
groups and did not preregister such a difference. Indeed, we find no
difference in likeability and warmth across the two treatments
(`r papaja::apa_print(t.test(likable ~ Condition, data = study2_time3, var.equal = T))$statistic`,
and
`r papaja::apa_print(t.test(warmAvg ~ Condition, data = study2_time3, var.equal = T))$statistic`,
respectively). We also assess the quality of advice by measuring the
discrepancy between the advised rank and the final rank. Our analysis
revealed no significant difference between the two treatments in terms
of how helpful the advice was for advisees in making accurate guesses,
`r papaja::apa_print(t.test(accuracy ~ Condition, data = study2_time3, var.equal = T))$statistic`.

@@

In another context where advice communicates ego-relevant information
(here, people's attractiveness), we find that the advice people give is
contingent on their incentives. Specifically, when they get rewarded for
being more likeable, they recommend that the advisee bet on a more
favorable rank than when they are incentivized for accuracy.
Importantly, advisees do not discount flattering advice and instead
evaluate people who advise them to bet on a more attractive rank as
warmer and more likeable. These gains to interpersonal perceptions do
not come at the cost of trustworthiness, even when the advice is
substantially inflated (e.g., a recommendation that the advisee is the
most attractive in the group).

>>>>

We next examine if advisers believe that flattering advice will improve
how they are perceived, and if so, whether this perception is true. We do this
by manipulating the incentives for advisers, who receive a bonus either based
on how they are evaluated by an advisee, or a bonus for the advisee's accuracy.

The study takes place in three stages. First, a group of
advisees upload photos of themselves ("selfies") and are grouped with 
nine other participants of the same sex We then recruit participants of
the opposing sex to rank them from most to least attractive and to
provide advice. Specifically, they advise the participant
they ranked as the 7th most attractive (i.e., 4th least attractive) on what
rank they should bet they were ranked by a larger group of raters.
Advisers were randomly assigned to two treatments, receiving a bonus
payment either if the advisee guessed their rank accurately or if the
advisee evaluated the adviser as likeable as measured by a scale response. 
We hypothesize that advisers incentivized to be liked will recommend betting on
a lower rank, i.e. that the advisee is more attractive.

@@

We recruit 300 participants from Prolific and, after asking
demographic questions, invite them to upload photos of themselves
(selfies) to be rated by other participants on attractiveness.
We obtain selfies from `r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women adhering to our
instructions (e.g., did not include other people). In line with our preregistration,
we select the first 100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years). Participants
are informed that their selfies will be randomly grouped with those of
nine other participants of their sex and ranked in terms of
attractiveness by a group of new Prolific participants of the opposite
sex. They then make an unincentivized guess of their rank.

Next, we recruit `r scales::comma(dim(study2_time2)[1])` participants from Prolific
for the role of advisers ($M_{\text{Age}}$ = `r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female). We first
collect demographic information, then match them to a group of
the opposite sex. They rank selfies from ten participants from most to least
attractive using a drop-down menu next to each picture. Because of a
limitation with the survey software, we could not validate that each rank is
given only once, and we remove 115 participants who failed to follow instructions 
and did not provide a complete ranking.^[Moreover, we could not collect data from any participant who did not select any participant as the 7th most attractive participant, as that image would be shown on subsequent pages.]

After submitting their ratings, participants see the photo of the participant
they ranked as the 7th most attractive (i.e., the 4th least
attractive). We remind them of the rank they have just given to that
person and inform them that this participant will be invited back and
can earn a \$1 bonus if they correctly guess their rank. The rank
is determined by the aggregate ratings of all participants who have
ranked this group. Because the participant does not observe the other nine
people in the group, they would depend on the adviser's recommendation along with
their own assessment. We randomly assign advisers to
one of two incentivization schemes. In the "Accuracy" treatment, they
receive a bonus identical to the advisee: \$1 if they guess their
rank correctly. In the "Likeability" treatment, we inform them that
the advisee will rate them on a 5-point Likert scale on how
likeable they thought they are. Each point on the scale would translate
to a bonus of 20 cents. Advisers then select a rank (from 1 to 10) that they
recommend the advisee to bet on.

Finally, we invite participants from Stage 1 who were ranked as the 7th most
attractive participant by at least one adviser (so that they received advice) back
for the follow-up survey. Following our preregistration, we keep the
survey open for 7 days. In total,
`r scales::comma(dim(study2_time3)[1])` participants
(`r scales::comma(dim(study2_time3[study2_time3$gender == 'Male',])[1])`
men,
`r scales::comma(dim(study2_time3[study2_time3$gender == 'Female',])[1])`
women) complete the follow-up survey. We remind them of the selfie they uploaded in Stage
1 and that a group of 10 selfies, including theirs, had been
rated by other participants from Prolific. They then see the advice from
a randomly selected adviser and make their own guess with a \$1 incentive
for accuracy. Finally, they rate the adviser's likability,
warmth, friendliness, good-naturedness, trustworthiness, and sincerity
on 5-point Likert scales [adapted from @Fiske2007].

@@

```{r}
#| label: fig-study2advice
#| fig-cap: "Advisers were asked to suggest what rank to bet on for someone they ranked as the 7th most attractive out of ten people. When incentivized for accuracy, most of them advised betting on 7, which was based on their evaluation of the recipient's attractiveness. However, when incentivized for likability, many participants gave flattering advice by suggesting betting on higher ranks that imply a higher evaluation of attractiveness."
#| eval: false

@@

    # Theme adjustments
    theme_bw() +
  coord_flip() +
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(colour = "black"),
          legend.position = "none",
          legend.title = element_blank(),
          legend.key.size = unit(1.2, "lines"),
          legend.text = element_text(size = 12),
          plot.title = element_text(hjust = 0.5, size = 20),
          axis.text.x = element_text(size = 14),  # X-axis text size
          axis.text.y = element_text(size = 14),
          axis.title.x = element_text(size = 16, face = "bold"),  # X-axis title size
          axis.title.y = element_text(size = 16, face = "bold"))

```
```{r}
#| label: fig-study2advice
#| fig-cap: "Participants in Study 2 give advice to the person they rank 7th most attractive out of 10. When incentivized for accuracy, the majority recommends betting on that rank. However, when incentivized for likeability, participants inflate their advice to flatter the recipient as more attractive."
study2_time2 %>%
  dplyr::group_by(Condition) %>%
  dplyr::mutate(total_per_condition = dplyr::n()) %>%
  dplyr::group_by(Condition, Advice, total_per_condition) %>%
  dplyr::summarize(n = dplyr::n()) %>%
  dplyr::mutate(pct = ifelse(Condition == "Accurate", -n/total_per_condition, n/total_per_condition),
                Advice = factor(Advice)) %>%
  ggplot(aes(x = pct, y = Advice, fill = Condition)) +
  geom_col() +
  scale_x_continuous("Share of Advice",
                     limits = c(-0.6, 0.6),
                     labels = \(x) scales::percent(abs(x))) +
  scale_y_discrete(limits = rev) +
  # Dashed 'horizontal' line at Advice = 7
  geom_hline(yintercept = 4, col = 'black', linetype = 'dashed', linewidth = 0.7) +
  # Example text annotation (adjust x/y for positioning)
  annotate("text", x = -0.3, y = 9, label = "Accuracy Incentive", hjust = 1, size = 4, fontface = "bold") +
  annotate("text", x = 0.6, y = 9, label = "Likeable Incentive", hjust = 1, size = 4, fontface = "bold") +
  # scale_y_continuous(name = "Advised Rank", breaks = seq(1, 10)) +
  scale_fill_manual(values = c("#2A629A", "#FF7F3E")) +
  papaja::theme_apa() +
  theme(
    legend.position  = "none",
    axis.text.x  = element_text(size = 14),
    axis.text.y  = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold")
  )
```

We begin by examining the prior beliefs of advisees who uploaded their
selfies. On average, men guess they rank
`r round(mean(study2_time1[study2_time1$gender == "Male",]$Guess),2)` in
their group of 10 and women guess that they rank
`r round(mean(study2_time1[study2_time1$gender == "Female",]$Guess),2)`
(`r papaja::apa_print(t.test(Guess ~ gender, data = study2_time1, var.equal = T))$statistic`).
Notably, people's self-perceptions correlate strongly with the
aggregate ratings of the advisers
(`r round(cor(study2_finalRank$Guess, study2_finalRank$finalRank),2)`,
`r papaja::apa_print(cor.test(study2_finalRank$Guess, study2_finalRank$finalRank))$statistic`).
However, there is substantial heterogeneity in perceptions of
attractiveness. Of the 200 participants,
`r length(unique(study2_time2$Target))` are ranked as 7th most
attractive by at least one adviser. On average, men in this subset
estimate they rank
`r round(mean(TargetsUnique[TargetsUnique$gender == 'Male',]$Guess),2)`th
and women estimate they rank
`r round(mean(TargetsUnique[TargetsUnique$gender == 'Female',]$Guess),2)`th
(`r papaja::apa_print(t.test(Guess ~ gender, data = TargetsUnique, var.equal = T))$statistic`).

Next, we turn our attention to the advisers (see @fig-study2advice). In the
Accuracy condition, those uploading selfies are on average advised to bet on rank
`r round(mean(study2_time2[study2_time2$Condition == "Accurate",]$Advice),2)`.
Notably, this is significantly more attractive than the 7th rank those
advisers had themselves guessed just on the prior screen
(`r papaja::apa_print(t.test(study2_time2[study2_time2$Condition == "Accurate",]$Advice, mu = 7, var.equal = T))$statistic`).
This suggests that even when incentivized for accuracy, participants
offer flattering advice.[^1] Importantly, and as predicted, we find
that advisers in the Likeable treatment recommend betting on a lower
rank, communicating that they think the participant in the selfie is
more attractive
(`r round(mean(study2_time2[study2_time2$Condition == "Likable",]$Advice),2)`,
`r papaja::apa_print(t.test(Advice ~ Condition, data = study2_time2, var.equal = T))$statistic`).
This suggests that advisers infer flattering someone with pleasant advice
would make the adviser appear more likeable. The distribution shown in
@fig-study2advice shows that participants do not simply tell
participants that they are the most attractive person in the group. They
may infer that flattering advise needs to be somewhat realistic to be
believable. We return to this in the general discussion.

[^1]: This shading could be due to concerns of avoiding disappointment.
    However, it could also be that advisers are uncertain about the
    rankings they have given and make a recommendation that combines
    their own belief with a uniform prior. Therefore, our analyses focus
    on the difference between the two conditions.

@@

# Create the plot
ggplot(study2_long, aes(x = Advice, y = Value, color = Evaluation)) +  
  # Add points for each observation
  geom_point(size = 2, alpha = 0.7,
             position = position_jitter()) +

@@

```{r}
#| label: fig-study2ratingByGroup
#| fig-cap: "After participants receive advice that suggests betting on a rank that was lower than, equal to, or higher than their own guess, they evaluate advisors based on their likability and warmth. Participants rate advisors who give flattering advice and deliver a high evaluation of their attractiveness as more likable and warm than those whose advice implied a lower evaluation of their attractiveness."

@@

# Create the heatmap
ggplot(melted_data, aes(x = Advice, y = Condition, fill = Rating)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "darkblue", high = "orange", name = "Rating") +  # Grey gradient from light to dark
  geom_text(aes(label = sprintf("%.2f", Rating)), color = "white", size = 8) +
  theme_minimal() +
  labs(x = "Attractiveness (Advice Relative to Self-Expectations)", y = "Evaluation") +
  theme(
    axis.text.x = element_text(size = 12),  # X-axis text size
          axis.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 12),  # X-axis title size
          axis.title.y = element_text(size = 12))

@@

As seen in @fig-study2rating, we advisers who suggested that
the advisee was more attractive (lower rank) were indeed rated as more
likeable and warm (b = -0.109, p \< .05; b = -0.133, p \< .001,
respectively; See Columns 1 and 2 of @tbl-study2regs). When using their
own initial guess as a reference point, advisers who suggest a worse rank
are perceived as less likable and warm (see
@fig-study2ratingByGroup). Interestingly, these benefits are not at the
cost of sincerity; advisers who recommend a more favorable rank
are viewed as no less trustworthy (Column 3 of @tbl-study2regs).

We are not powered to do a comparison across the two experimental
groups and did not preregister such a difference. Indeed, we find no
difference in likeability and warmth across the two treatments
(`r papaja::apa_print(t.test(likable ~ Condition, data = study2_time3, var.equal = T))$statistic`,
and
`r papaja::apa_print(t.test(warmAvg ~ Condition, data = study2_time3, var.equal = T))$statistic`,
respectively). We also assess the quality of advice by measuring the
discrepancy between the advised rank and the correct rank. We find
no difference in accuracy across the advice treatments
(`r papaja::apa_print(t.test(accuracy ~ Condition, data = study2_time3, var.equal = T))$statistic`).

@@

In another context where advice communicates ego-relevant information
(here, people's attractiveness), we find that the advice people give is
contingent on their incentives. Specifically, when they get rewarded for
being more likeable, they recommend that the advisee bet on a more
favorable rank than when they are incentivized for accuracy.
Importantly, advisees do not discount flattering advice and instead
evaluate people who advise them to bet on a more attractive rank as
warmer and more likeable. These gains to interpersonal perceptions do
not come at the cost of trustworthiness, even when the advice is
inflated.
